Open Clusters Characterization in Gaia DR2 Using ML Algorithms

### 2020-11-11 thesis/thesis.tex
thesis: Add Data Mining section

The development has been made with the Python programming language. Other auxiliary tools such as \emph{Docker}, \emph{PostgreSQL}, \emph{Jupyter Notebooks},
and \emph{Scikit-Learn} and \emph{Keras} frameworks have been used too.

@@

As mentioned before, Gaia DR2 is the chosen data source for this work. Thus, the first step is to download the required data from Gaia DB
and store it in a custom database for later access.

@@

There is no upper limit to the number of stars inside a cluster, although for the sake of this work, smaller cluster will be prefered as
a commitment to the project delivery dates and the available computing power. As an extension of this work, the trained model could be applied to a region
not covered by this preselection for a later understanding of the results.

>>>>

All code is developed with the Python programming language \cite{Python3}. Other auxiliary tools such as \emph{Docker} \cite{merkel2014docker},
\emph{PostgreSQL} \cite{postgresql}, \emph{Jupyter Notebooks} \cite{Kluyver2016jupyter}, and \emph{Astropy} \cite{astropy:2013} \cite{astropy:2018},
\emph{Scikit-Learn} \cite{scikit-learn}, \emph{seaborn} \cite{michael_waskom_2017_883859}, \emph{SQLAlchemy} \cite{sqlalchemy} and
\emph{Keras} \cite{chollet2015keras} frameworks have been used too.

@@

As mentioned before, I have chosen the Gaia DR2 database as the data source for this work. Thus, the first step is to download the required data from Gaia DB
and store it in a custom database for later access.

@@

As shown in Figure \ref{fig:OpenClustSelection} this selection stills being a good representation of all clusters in the Milky Way since they are
equally distributed around the galaxy line.

After this selection, the number of cluster to look up is 169 with nearly 75 million stars.

Since there is no upper limit to the number of stars inside a cluster, for the sake of simplicity and as a commitment to the project delivery dates
and the available computing power, smaller clusters will be prefered over the greatest ones.

As an extension of this work, the designed model could be applied to a region not covered by this preselection for a later understanding of the results.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{../figures/cluster_selection_tier1.pdf}
  \caption{OpenClust Catalogue Selection Distribution}
  \label{fig:OpenClustSelection}
\end{figure}

The download process has been performed with two Docker containers: a \emph{downloader} and the \emph{database}.

The first one is a container built from a \verb|python:3.8| image that contains the \verb|cdalvaro| package and the \verb|downloader.py| script.
This script is prepared to load the OpenClust catalogue, connect to the Gaia DR2 database, download all stars contained inside each cluster
(with a radius multiplier to increse the area to be downloaded) and save the downloaded data inside the custom database hosted in the second container.

The second container is just a \verb|postgres:12.4| image which loads an initial script when the database is not yet initialized for creating the
database structure. This container is used later as the main database for data analysis.

The database has two main tables, \verb|regions| for storing cluster properties such as location, diameter and other properties, and \verb|gaiadr2_source|
which contains data for all downloaded stars. This second table is partitioned by region, storing all stars related to one cluster inside an independent
table for faster access.

These containers are orchestrated with a \verb|docker-compose.yml| file which automatilly launches one instance of each container.

The code developed for this work can be found inside the \verb|src| directory at GitHub:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis}
